---
title: "Introduction to quick.lctm.escc"
author: "Anonymous"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to quick.lctm.escc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(quick.lctm.escc)
```

## What is quick.lctm.escc?

quick.lctm.escc fits a Bayesian Latent Class Trajectory Model (LCTM) on **7
nutritional / inflammatory markers** collected at up to 5 time points
during neoadjuvant chemotherapy for oesophageal cancer.  The model
classifies each patient into one of two latent classes:

* **Resilient Class (RC)** -- nutritional trajectory stays high.
* **Vulnerable Class (VC)** -- nutritional trajectory drops
  sharply, indicating a need for intensive prehabilitation.

The package uses only the 7 markers (ALB, TC, LYM, CRP, SMI, SFI, VFI);
it does not require baseline demographic variables (sex, BMI, age,
etc.) and ships a pre-trained v8 posterior for offline classification
without re-running MCMC.

## Required input format

A long-format `data.frame` with one row per (patient, time-point):

| Column | Type | Example values |
|--------|------|----------------|
| `id` | integer / character | `550193` |
| `timepoint` | character | `T1` / `T2` / `T4` / `T5` / `T6` (or `T1_pre`, `T2_post`, `T4_3M`, `T5_6M`, `T6_1Y`) |
| `ALB` | numeric | `43.5` g/L |
| `TC` | numeric | `4.95` mmol/L |
| `LYM` | numeric | `1.85` 10^9/L |
| `CRP` | numeric | `7.4` mg/L |
| `SMI` | numeric | `33.9` cm^2/m^2 (skeletal muscle index, CT) |
| `SFI` | numeric | `62.2` cm^2/m^2 (subcutaneous fat index, CT) |
| `VFI` | numeric | `60.4` cm^2/m^2 (visceral fat index, CT) |

At least 2 distinct time-points per patient are required.

## The 6-step user workflow

### 1. Install

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Tanly78/quick.lctm.escc")
```

### 2. Load the package and prepare your data

```{r}
library(quick.lctm.escc)
patients <- read.csv(
  system.file("extdata", "example_lctm.csv", package = "quick.lctm.escc"),
  stringsAsFactors = FALSE
)
str(patients)
```

### 3. Classify with the pre-trained v8 posterior

```{r}
pred <- predict_lctm7m(patients, use_pretrained = TRUE)
head(pred)
```

`predict_lctm7m()` returns a `data.frame` with columns
`id`, `p_VC`, and `cluster` (factor with levels `RC` < `VC`).
Wall-clock time is < 0.1 s for 1000 patients because no MCMC is
involved -- the function applies the v8 posterior means already bundled
in the package.

### 4. Stratify

```{r}
table(pred$cluster)
```

### 5. (Optional) Compute the Trajectory Bend Score

```{r}
t1 <- patients[patients$timepoint %in% c("T1", "T1_pre"), ]
t2 <- patients[patients$timepoint %in% c("T2", "T2_post"), ]
tbs <- trajectory_bend_score(t1, t2)
head(tbs)
```

### 6. (Optional) Export results

```{r, eval = FALSE}
write.csv(pred, "patient_classification.csv", row.names = FALSE)
```

## Verifying the v8 posterior on your own cohort

The v8 posterior is reproducible from the maintainer-hosted
`stan_lctm_v8_K2_fit.rds` (41 MB, not shipped).  Use
`data-raw/extract_v8_params.R` to regenerate the bundled
`lctm_v8_params` rda.

## Next steps

* `vignette("L1_full_analysis")` -- train a fresh LCTM on your own
  cohort using `fit_lctm_full()` (requires `rstanarm`).
* `?predict_lctm7m` -- full argument reference.
* `?lctm_v8_params` -- details of the bundled posterior.